### 2012-10-15 thesis/tex/3_background.tex
initial commit v2 ;)

In the beginning all programming of the robot behavior was done by directly editing the XML code mentioned in the previous section by using a text editor. But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise. For this Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behavior on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behavior on a visual layer and generates the corresponding program file containing the RBDL code fully automatically. This file can be downloaded to the robot and executed by the interpreter. Figure~\ref{fig:robostudio_proceeding} illustrates what the steps of robot behavior development are.

>>>>

In the beginning all programming of the robot behavior was done by directly editing the XML code mentioned in the previous section by using a text editor. But huge code files and occasional code edits made further maintenance and change requests difficult up to impossible to realise. For this reason Chandan Datta, PhD student at the University of Auckland, developed a visual programming environment for developing robot behavior on top of RBDL: Robostudio~\cite{robostudio} allows to easily edit the program behavior on a visual layer and generates the corresponding program file containing the RBDL code fully automatically. This file can be downloaded to the robot and executed by the interpreter. Figure~\ref{fig:robostudio_proceeding} illustrates what the steps of robot behavior development are.

### 2012-10-15 thesis/tex/5_design.tex
initial commit v2 ;)

One main difficulty in finding a proper visual formalism was to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points. A first approach, so called template constraints, and its disadvantages are described in section~\ref{sec:templateconstraints}. Subsequently an improved design is presented, which tends to a more abstract direction.

@@

Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in section~\ref{sec:automatedconstraintgeneration} shall be assembled in one tool, the LTLCreator. It shall provide a dashboart with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results. Besides a button is needed for triggering constraint generation.
The LTLCreator should be available for other state machine based programming environments for integration.
The implementation as a java swing component and a clear API enables the tool for a wide range of use.

Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be. The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program. It will automatically revalidate all existing constraints and provide the up-to-date model for constraint generation.
For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.

@@

TODO: NuSMV hier auch wichtig?

@@

Robostudio uses NetBeans as a platform, and all tools and features are organized in windows. Thus also the LTLCreator has to be put into a window and added to the Robostudio platform. Since the defining of safety constraints is an important part while developing service robot behavior, the LTLCreator window shall appear in the center of the platform.

>>>>

One main difficulty in finding a proper visual formalism is to glean the right level between abstraction and expressiveness since this two characteristics appeared to be mutually exclusive in some points. A first approach, so called template constraints, and its disadvantages are described in section~\ref{sec:templateconstraints}. Subsequently an improved design is presented, which tends to a more abstract direction.

@@

Both features the visual language presented in section~\ref{sec:operatorconstraints} and the automated constraint generation in section~\ref{sec:automatedconstraintgeneration} shall be assembled in one tool, the LTLCreator. It shall provide a dashboart with the possibility to manage several constraints, such as a tab panel, and displays for indicating validation results. Besides a button is needed for triggering constraint generation.
LTLCreator should be available for other state machine based programming environments for integration.
The implementation as a java swing component and a clear API enables the tool for a wide range of use.

Figure~\ref{fig:coalescence} points out how interaction between LTLCreator and programming environments where it is integrated in can be.
LTLCreator consists of a dashboard where constraints can be edited visually and an graphical output of validation results. The constraint generator as well as the validator are non UI parts of LTLCreator.
The programming environment provides the state machine model and notifies the LTLCreator component about all changes made to the program. This will automatically revalidate all existing constraints and always provide the up-to-date model for constraint generation.
For load and save functionality, all existing constraints can be received and new ones can be added to the LTLCreator component programmatically.

@@

@@

Robostudio uses NetBeans as a platform, and all tools and features are organized in windows. Thus also LTLCreator has to be put into a window and added to the Robostudio platform. Since the defining of safety constraints is an important part not only after but also during the process of developing service robot behavior, the LTLCreator window as a vital tool shall be placed in the center of the platform where currently also the UI Component Layout Editor Window is. As already proposed in the previous section, all changes to the state machine program made within robostudio must of course cause the LTLCreator window to revalidate all existing constraints.

### 2012-10-15 thesis/tex/6_implementation.tex
initial commit v2 ;)

\chapter{The LTLCreator prototype}
\section{Operator constraints}
\section{Automated constraint generation}
\section{Model checker}

@@

>>>>

\chapter{The LTLCreator prototype}

As stated in chapter~\ref{chap:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language. We therefore designed and implemented a prototype of such an editor. The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated. The implementation is a java swing component, making it easy to integrate the visual language into other java applications. An API enables an underlying state machine which also influences the state propositions available for constraint editing. Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.

Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed in the dashboard in the center of the editor by drag\&drop. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{editor} 
  \caption{Snapshot of the visual editor.}
  \label{fig:editor}
\end{figure}

TODO: Kleine Inhaltsangabe

\section{Operator constraints}
\section{Automated constraint generation}

% (on huge state machines) (brute force algorithm)

%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde, der saemtliche sub graphs durch Abwandern sämtlicher Pfade der zugrundeliegenden state machine herauszufinden versucht, kann bei grossen programmen (ab ca. 1000 Zustaenden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern. Um den warteprozess ertraelicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.

\section{Model checker}

@@

statische analyse

### 2012-10-15 thesis/tex/7_prototypeandevaluation.tex
initial commit v2 ;)

case study, eventuell grosses Beispiel erwaehnen? wenn constraints generiert werden, verlaesst man sich darauf? Hoert man dann auf, selbst mitzudenken? Ist die menge der constraints noch handelbar?

As stated in Sect.~\ref{sec:goals}, appropriate support by a graphical editor is important for the usefulness of the developed visual language. We therefore designed and implemented a prototype of such an editor. The proposed constraint generation heuristics is integrated in this prototype as well and could thus be evaluated. The implementation is a java swing component, making it easy to integrate the visual language into other java applications. An API enables an underlying state machine which also influences the state propositions available for constraint editing. Furthermore a particular model checker can be specified for constraint validation. As a default, the symbolic model checker NuSMV~\cite{springerlink:10.1007/s100090050046,NuSMV2} is used.

Fig.~\ref{fig:editor} shows a snapshot of the editor's environment: All functionality needed for constraint editing is provided in the tool bar on the right. It contains draggable elements for creating all operator and proposition types as well as a trashcan for deleting. Constraints can be composed in the dashboard in the center of the editor by drag\&drop. The tab functionality on the left allows multiple constraints to be managed. Each tab shows a small thumbnail of the constraint and a symbol indicating its validity (a warning shield for ``incomplete'', a green shield for ``valid'', a red cross for ``invalid'' or an animated ring for ``validation in progress.'') The magic wand button within the tab pane triggers the constraint generation.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{editor} 
  \caption{Snapshot of the visual editor.}
  \label{fig:editor}
\end{figure}

In order to evaluate the proposed safety functionality the tool was integrated into Robostudio~\cite{robostudio}. This Visual Programming Environment allows the specification of state machine based programs whose behaviour can then be examined by visual constraints. As an example we consider the simplified medication reminder application whose underlying screen-flow is explained in Fig.~\ref{fig:medicationreminder}. Starting from ``Menu'' the user can choose from several services such as blood pressure measurement or entertainment. In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part. In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered. The patient can state if he or she has already taken the medication, or decide whether to take it or not. In the latter he or she may give a reason for it and a caregiver is notified about this incident. Otherwise it will result in ``Well done!'' after completing intake.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{stm.pdf} %width=0.7\linewidth
  \caption{Simplified screen-flow of the medication reminder application.}
  \label{fig:medicationreminder}
\end{figure}

@@

Before we can start creating constraints it might be useful to think about a reasonable term we want to express. Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given in the requirements in section~\ref{sec:requirements}.

@@

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{example_propositions}
  \caption{Possible proposition extensions: numeric and string equation.}
  \label{fig:example_propositions}
\end{figure}

\section{Automated Constraint generation}

The automated constraint generation can be triggered by a click on the magic wand button in the tab area. After activation a dashboard is opened in a new tab for each found constraint, and validation is initiated immediately. For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in Fig.~\ref{fig:generatedconstraints} which match the postulations in the requirements in Sect.~\ref{sec:requirements}.
The constraint used for demonstration in the previous section is also generated, however \emph{b)} forms an intensified restriction of it.

Constraint \emph{a)} ensures the ``Polling'' state is always eventually visited again. If a pending medication is not already taken, constraint \emph{b)} guarantees intake or staff notification before the next reminders can be read from the database.

@@

%** COMMENT: Possible comment here about the scalability and practicality about constraint generation; do you think it will explode in larger cases? Either way a comment might be useful. **

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{generatedconstraints} %width=0.7\linewidth
  \caption{Automatically generated constraints which match the postulations in the requirements.}
  \label{fig:generatedconstraints}
\end{figure}

%Da der subgraph-finde-algorithmus als brute force algorithmus implementiert wurde kann bei groï¿½en programmen (ab ca. 1000 Zustï¿½nden und sehr vielen Verzweigungen) der findeprozess teils sehr lange (mehrere Minuten) dauern. Um den warteprozess ertrï¿½glicher zu machen wird wï¿½hrend der findung ein dialog mit dem aktuellen fortschritt und einem abbrechen button angezeigt.
% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht

@@

 (on huge state machines) (brute force algorithm)

@@

@@

\section{Reliability}

>>>>

In order to evaluate the proposed safety functionality we use the Robostudio version where we already integrated LTLCreator in. This visual programming environment allows the specification of state machine based programs whose behaviour can then be examined by visual constraints.
As a test scenario we consider the simplified medication reminder application whose underlying screen-flow is explained in figure~\ref{fig:medicationreminder}. Starting from ``Menu'' the user can choose from several services such as blood pressure measurement or entertainment. In the given example these services are simplified to just one state ``additional services'' since we want to focus only on the medication reminder part. In ``Polling'' the database is checked for a pending reminder, and if there is one the medication intake guidance is triggered. The patient can state if he or she has already taken the medication, or decide whether to take it or not. In the latter he or she may give a reason for it and a caregiver is notified about this incident. Otherwise it will result in ``Well done!'' after completing intake.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{stm.pdf}
  \caption{Simplified screen-flow of the medication reminder application.}
  \label{fig:medicationreminder}
\end{figure}

This state machine with all its states, transitions and screen dialogs has been implemented in Robostudio and in the following sections we will evaluate the visual language as well as the capability of constraint generation. In section~\ref{sec:reliability} a short statement about reliability of the features is given.

@@

Before we can start creating constraints it might be useful to think about a reasonable term we want to express. Let's take the example ``Whenever there is a pending reminder, medication will be finally taken or caregivers will get notified in case of the patient refusing medication intake.'' given as a requirement in chapter~\ref{chap:goals}.

@@

We observed the different color flavours of the visual operators to be a good support for fast reading and understanding of constraints. Also the two dimensional composition and the round shapes of the operators turned out to 
give the language a schematic but not too rectangular look. Besides it is considered eye-candy what is the best motivation for using this visual language.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{example_propositions}
  \caption{Possible proposition extensions: numeric and string equations.}
  \label{fig:example_propositions}
\end{figure}

\section{Automated constraint generation}

The automated constraint generation can be triggered by a click on the magic wand button in the tab area. After activation a dashboard is opened in a new tab for each found constraint, and validation is initiated immediately. For the medication reminder application six constraints are found, including \emph{a)} and \emph{b)} shown in figure~\ref{fig:generatedconstraints} which match the postulations in the requirements in chapter~\ref{chap:goals}.
The constraint used for demonstration in the previous section is also generated, however \emph{b)} forms an intensified restriction of it. 
Constraint \emph{a)} ensures the ``Polling'' state is always eventually visited again. If a pending medication is not already taken, constraint \emph{b)} guarantees intake or staff notification before the next reminders can be read from the database.

The search and generation process finishes within less than a second and thus doesn't make the developer wait a long time.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.65]{generatedconstraints}
  \caption{Automatically generated constraints which match the required postulations.}
  \label{fig:generatedconstraints}
\end{figure}

@@

In order to examine the performance of sub graph finding and constraint generation, we need to have a second look at the implementation details again. The sub graph finding algorithm is realized as a depth-first search. Indeed all healthcare programs are infinite and contain cycles and a depth-first walk-through would never end, but the algorithm doesn't go deeper at already visited states.

The performance of this algorithm is of course highly dependend on the grade of branching. Nevertheless, such a depth-first search has an estimated wost case execution time of $O(|V|+|E|)$ in which $|V|$ is the number of states and $|E|$ number of edges respective transitions in our case.
Since we cope with state machines and thus have directed edges rather than undirected ones, each state can have $|E|=|V|$ transitions; $|V-1|$ to other ones and one to itself. As worst case ececution time for the depth-first search we obtain the following result:

\begin{equation}
O(\textnormal{``depth-first search''})=O(|V|)+O(|V|-1)=O(|V|)
\end{equation}

In order to find a sub graph with our presented algorithm the depth-first search has to be startet on a start state of a sub graph. Since we don't know in beforehand which states might be possible candidates for such start states, the algorithm has to be applied on every single state of the program which has two ore more outgoing transitions. In the worst case all $|V|$ states match this condition, so the depth-first search has to be executed $|V|$ times. This implies for our estimated worst case execution time:

\begin{equation}
O(\textnormal{``sub graph finding''})=O(|V|)*O(|V|)=O(|V|^2)
\end{equation}

@@

For all the healthcare applications the presented algorithm works brilliant and speedy. But their program sizes are still quite manageable.
Of course an exponential execution time is bad if the number of states arises. A clear slowdown can be felt when the algorithm runs on larger state machines with plenty of states and high grade of branching.

However, sholud this algorithm regardless its tailoring to the healthcare domain be used for other huge programs, it keeps a lot of potential to be improved in efficiency. With a sophisticated approach it might be enough to execute just one depth-first search for finding all sub graphs at once what would lead to a linear execution time.
% wenn noch schreiben will, dann: linear ist cool, weil das tool eh die gefundenen constraints gleich anzeigt und nicht erst am ende alles geballt bringt.

@@

% hat uns die automatische generierung geholfen? Wuhu, wir haben noch einen constraint gefunden, der auch sinn macht
%practicality of constraints

\section{Reliability}
\label{sec:reliability}
